---
title: 'STAT 429 Project: US Housing Market'
author: "Kunal Bhardwaj"
date: "2024-02-23"
output:
  html_document: default
  pdf_document: default
---

The objective is to predict the Median Sales price of houses sold in the United States based on factors like:

i) Monthly supply of New Houses in the United States
ii) New Privately-owned Housing units started
iii) Home ownership rate in the United States
iv) 30-Year Fixed Rate Mortgage Average in the United States

The Median Sales Price will be adjusted based on Consumer Price Index (CPI) for all Urban Consumers: All items less shelter. This will give us an idea of how much house prices have risen in excess of inflation in other commodities.


```{r message=FALSE, warning=FALSE}
library(readr)
MSPUS <- read_csv("Data/MSPUS.csv",
                  col_types = cols(DATE = col_date(format = "%Y-%m-%d")))

CUSR0000SA0L2 <- read_csv("Data/CUSR0000SA0L2.csv",
                          col_types = cols(DATE = col_date(format = "%Y-%m-%d"),
                                           CUSR0000SA0L2 = col_double()))

MSACSR <- read_csv("Data/MSACSR.csv",
                   col_types = cols(DATE = col_date(format = "%Y-%m-%d")))

HOUST <- read_csv("Data/HOUST.csv",
                  col_types = cols(DATE = col_date(format = "%Y-%m-%d"),
                                   HOUST = col_number()))

RHORUSQ156N <- read_csv("Data/RHORUSQ156N.csv", 
    col_types = cols(DATE = col_date(format = "%Y-%m-%d")))

MORTGAGE30US <- read_csv("Data/MORTGAGE30US.csv",
                         col_types = cols(DATE = col_date(format = "%Y-%m-%d"),
                                          MORTGAGE30US = col_number()))
```


```{r message=FALSE}
library(astsa)
library(ggplot2)
library(ggfortify)
library(fpp2)
library(tseries) # To carry out ADF & KPSS Tests
library(lmtest) # To carry out BP-Test

mspus.ts <- ts(MSPUS$MSPUS[34:244], start = c(1971, 2), frequency = 4)
cusr.ts <- ts(CUSR0000SA0L2$CUSR0000SA0L2[98:308], start = c(1971, 2), frequency = 4)

msacsr.ts <- ts(MSACSR$MSACSR[34:244], start = c(1971, 2), frequency = 4)
houst.ts <- ts(HOUST$HOUST[50:261], start = c(1971, 2), frequency = 4, end = c(2023, 4))
rhorusq156n.ts <- ts(RHORUSQ156N$RHORUSQ156N[26:236], start = c(1971, 2), frequency = 4)
mortgage.ts <- ts(MORTGAGE30US$MORTGAGE30US, start = c(1971, 2) , frequency = 4, end = c(2023, 4))


tsplot(mspus.ts, col ="darkorange",
       main = "Median Sales Price of Houses Sold for the United States",
       ylab = "USD")

mspus.ts.adj <- mspus.ts/cusr.ts*100
lines(mspus.ts.adj, col = "mediumblue")
legend("topleft", legend = c("Real Prices", "Adjusted for Inflation"), col = c("darkorange", "mediumblue"), lty = 1)
```

From the plot of Median Sales Price of Houses Sold, we can see that there exists an obvious trend in the data. Therefore de-trending needs to be carried out to achieve stationarity.


### Monthly Supply of New Houses in the United States (MSACSR) [Quarterly Adjusted]

The months' supply is the ratio of new houses for sale to new houses sold. This statistic provides an indication of the size of the new for-sale inventory in relation to the number of new houses currently being sold. The months' supply indicates how long the current new for-sale inventory would last given the current sales rate if no additional new houses were built.

### New Privately-Owned Housing Units Started: Total Units (HOUST) [Quarterly Adjusted]

As provided by the Census, start occurs when excavation begins for the footings or foundation of a building. Increases in housing starts and permits indicate growing supply, which can help alleviate housing shortages and moderate price growth. Conversely, declines in construction activity may contribute to supply constraints and upward pressure on prices.

### Homeownership Rate in the United States (RHORUSQ156N)

The homeownership rate is the proportion of households that is owner-occupied.

### 30-Year Fixed Rate Mortgage Average in the United States (MORTGAGE30US) [Quarterly Adjusted]

On November 17, 2022, Freddie Mac changed the methodology of the Primary Mortgage Market Survey® (PMMS®). The weekly mortgage rate is now based on applications submitted to Freddie Mac from lenders across the country.

```{r message=FALSE}
par(mfrow=c(4,1))
tsplot(msacsr.ts, col ="mediumblue",
       main = "Monthly Supply of New Houses in the United States",
       ylab = "Month's Supply")
tsplot(houst.ts, col ="limegreen",
       main = "New Privately-Owned Housing Units Started : Total Units",
       ylab = "Thousands of Units")
tsplot(rhorusq156n.ts, col ="red",
       main = "Homeownership Rate in the United States",
       ylab = "Percent")
tsplot(mortgage.ts, col ="purple",
       main = "30-Year Fixed Rate Mortgage Average in the United States", ylab = "Percent")
```

```{r warning=FALSE}
# De-trending MSPUS Time Series
mspus_trend <- lm(mspus.ts.adj ~ time(mspus.ts.adj))

par(mfrow=c(2,1))
tsplot(mspus.ts.adj, col ="darkblue",
       main = "Median Sales Price of Houses Sold for the United States [Adjusted]",
       ylab = "USD")
abline(mspus_trend, col = "red")
tsplot(resid(mspus_trend), col ="darkblue",
       main = "[De-trended] Median Sales Price of Houses Sold for the United States",
       ylab = "Residual")

# check for Stationarity using (Augmented) Dickey-Fuller Test and KPSS Test
adf.test(resid(mspus_trend))
kpss.test(resid(mspus_trend))
bptest(lm(resid(mspus_trend) ~ time(resid(mspus_trend))))


mspus.ts.boxcox <- BoxCox(mspus.ts.adj, lambda = -0.35)
mspus.ts.boxcox.resid <- resid(lm(mspus.ts.boxcox ~ time(mspus.ts.boxcox)))
tsplot(mspus.ts.boxcox.resid)
adf.test(mspus.ts.boxcox.resid)
kpss.test(mspus.ts.boxcox.resid)
bptest(lm(mspus.ts.boxcox.resid ~ time(mspus.ts.boxcox.resid)))
```

ADF & KPSS tests conclude stationarity.

```{r warning=FALSE}
# Differencing MSPUS Time Series
tsplot(diff(mspus.ts.adj))

adf.test(diff(mspus.ts.adj))
kpss.test(diff(mspus.ts.adj))
bptest(lm(diff(mspus.ts.adj) ~ time(diff(mspus.ts.adj))))
```

The series passes both these tests of stationarity. But the series exhibits an obvious heteroscedasticity where higher levels are associated with higher variation. A log-transformation is recommended.

```{r}
# Log Transformation
tsplot(log(mspus.ts))
adf.test(log(mspus.ts))
kpss.test(log(mspus.ts))
bptest(lm(log(mspus.ts) ~ time(log(mspus.ts))))


# Return : Differencing after log transformation
tsplot(diff(log(mspus.ts)), main = "Return : Differencing after log transformation")
adf.test(diff(log(mspus.ts)))
kpss.test(diff(log(mspus.ts)))
bptest(lm(diff(log(mspus.ts)) ~ time(diff(log(mspus.ts)))))
```

The return (differencing after log transformation) of MSPUS failed both ADF and KPSS tests of stationarity and BP test of homoscedasticity.



The BP Test for homoscedasticity fails for MSPUS, log(MSPUS), diff(MSPUS), diff(MSPUS, lag =4),  diff(log(MSPUS)), and diff(resid(mspus_trend)). Will try differencing with a lag of 4 after log transformation.

```{r warning=FALSE}
# Differencing with lag = 4
tsplot(diff(mspus.ts, lag = 4))
adf.test(diff(mspus.ts, lag = 4))
kpss.test(diff(mspus.ts, lag = 4))
bptest(lm(diff(mspus.ts, lag = 4) ~ time(diff(mspus.ts, lag = 4))))

# Differencing with a lag of 4 after taking log
mspus.ts_log4 <- diff(log(mspus.ts), lag = 4)
tsplot(mspus.ts_log4)
adf.test(mspus.ts_log4)
kpss.test(mspus.ts_log4)
bptest(lm(mspus.ts_log4 ~ time(mspus.ts_log4)))

# Log transformation of the differencing at lag 4 is not carried out because it produces NaNs
```


```{r}
par(mfrow=c(3,2))
tsplot(log(mspus.ts), main = "Log Transformation: log(MSPUS)") # Log Transformation
tsplot(diff(mspus.ts), main = "Differencing with lag = 1") # Differencing
tsplot(diff(log(mspus.ts)), main = "Return : Differencing after log transformation") # Return (Differencing after log transformation)
tsplot(diff(residuals(mspus_trend)), main = "Differencing after de-trending") # Differencing after de-trending
tsplot(diff(mspus.ts, lag = 4), main = "Differencing with lag = 4 (Annual)") # Differencing with a lag of 4
tsplot(diff(log(mspus.ts), lag = 4), main = "Differencing with lag = 4 after log transformation")
```

```{r}
# ACF of transformed outcome variable
acf(diff(mspus.ts))
```


```{r}
tsplot(msacsr.ts, col ="mediumblue",
       main = "Monthly Supply of New Houses in the United States",
       ylab = "Month's Supply")
adf.test(msacsr.ts)
kpss.test(msacsr.ts)
bptest(lm(msacsr.ts ~ time(msacsr.ts)))
```

Thus the MSACSR series is stationary and homoscedastic.


```{r}
tsplot(houst.ts, col ="limegreen",
       main = "New Privately-Owned Housing Units Started : Total Units",
       ylab = "Thousands of Units")
adf.test(houst.ts)
kpss.test(houst.ts)
bptest(lm(houst.ts ~ time(houst.ts)))

# Differencing
houst.ts_diff <- diff(houst.ts)
tsplot(houst.ts_diff)
adf.test(houst.ts_diff)
kpss.test(houst.ts_diff)
bptest(lm(houst.ts_diff ~ time(houst.ts_diff)))
```

The ADF Test confirms stationarity of HOUST but KPSS does not. BP Test confirms homoscedasticity.

```{r}
tsplot(rhorusq156n.ts, col ="red",
       main = "Homeownership Rate in the United States",
       ylab = "Percent")
adf.test(rhorusq156n.ts)
kpss.test(rhorusq156n.ts)

rhorusq156n.ts_diff <- diff(rhorusq156n.ts)
tsplot(rhorusq156n.ts_diff)
adf.test(rhorusq156n.ts_diff)
kpss.test(rhorusq156n.ts_diff)
bptest(lm(rhorusq156n.ts_diff ~ time(rhorusq156n.ts_diff)))
```

Thus Home Ownership series passes the tests for stationarity and homoscedasticity after differencing.

```{r}
tsplot(mortgage.ts, col ="grey",
       main = "30-Year Fixed Rate Mortgage Average in the United States", ylab = "Percent")
adf.test(mortgage.ts)
kpss.test(mortgage.ts)

mortgage.ts_diff <- diff(mortgage.ts)
tsplot(mortgage.ts_diff)
adf.test(mortgage.ts_diff)
kpss.test(mortgage.ts_diff)
bptest(lm(mortgage.ts_diff ~ time(mortgage.ts_diff)))
```



```{r}
# Clipping the series to bring them to the same length
msacsr.ts <- ts(MSACSR$MSACSR[35:244], start = c(1971, 3), frequency = 4)
```


```{r}
trend <- time(diff(mspus.ts))
reg_model <- lm(diff(mspus.ts) ~ trend + msacsr.ts + houst.ts_diff + mortgage.ts_diff + rhorusq156n.ts_diff)
summary(reg_model)
summary(aov(reg_model))
```

