---
title: 'STAT 429 Project: US Housing Market'
author: "Kunal Bhardwaj"
date: "2024-02-23"
output:
  pdf_document: default
  html_document: default
---

```{r message=FALSE, warning=FALSE}
library(readr)
MSPUS <- read_csv("Data/MSPUS.csv",
                  col_types = cols(DATE = col_date(format = "%Y-%m-%d")))

MSACSR <- read_csv("Data/MSACSR.csv",
                   col_types = cols(DATE = col_date(format = "%Y-%m-%d")))

HOUST <- read_csv("Data/HOUST.csv",
                  col_types = cols(DATE = col_date(format = "%Y-%m-%d"),
                                   HOUST = col_number()))

RHORUSQ156N <- read_csv("Data/RHORUSQ156N.csv", 
    col_types = cols(DATE = col_date(format = "%Y-%m-%d")))

MORTGAGE30US <- read_csv("Data/MORTGAGE30US.csv",
                         col_types = cols(DATE = col_date(format = "%Y-%m-%d"),
                                          MORTGAGE30US = col_number()))
```


```{r message=FALSE}
library(astsa)
library(ggplot2)
library(ggfortify)
library(fpp2)
library(imputeTS) # For handling missing values by imputation
library(tseries) # To carry out ADF & KPSS Tests
library(lmtest) # To carry out BP-Test

mspus.ts <- ts(MSPUS$MSPUS[34:244], start = c(1971, 2), frequency = 4, end = c(2023, 4))
msacsr.ts <- ts(MSACSR$MSACSR[34:244], start = c(1971, 2), frequency = 4, end = c(2023, 4))
houst.ts <- ts(HOUST$HOUST[50:261], start = c(1971, 2), frequency = 4, end = c(2023, 4))
rhorusq156n.ts <- ts(RHORUSQ156N$RHORUSQ156N[26:236], start = c(1971, 2), frequency = 4, end = c(2023, 4))


# MORTGAGE30US contains missing values. They are occupied by "."
# They will be imputed by replacing with NA follwed by linear interpolation
#MORTGAGE30US$MORTGAGE30US <- ifelse(MORTGAGE30US$MORTGAGE30US == ".",
#                                    NA, MORTGAGE30US$MORTGAGE30US)
#MORTGAGE30US$MORTGAGE30US <- na_interpolation(as.numeric(MORTGAGE30US$MORTGAGE30US))
# Select 2nd column from MORTGAGE30US. The first value on Apr 2 1971 falls on the 14th week
mortgage.ts <- ts(MORTGAGE30US$MORTGAGE30US, start = c(1971, 2) , frequency = 4, end = c(2023, 4))

par(mfrow=c(5,1))
tsplot(mspus.ts, col ="darkorange",
       main = "Median Sales Price of Houses Sold for the United States",
       ylab = "USD")
tsplot(msacsr.ts, col ="mediumblue",
       main = "Monthly Supply of New Houses in the United States",
       ylab = "Month's Supply")
tsplot(houst.ts, col ="limegreen",
       main = "New Privately-Owned Housing Units Started : Total Units",
       ylab = "Thousands of Units")
tsplot(rhorusq156n.ts, col ="red",
       main = "Homeownership Rate in the United States",
       ylab = "Percent")
tsplot(mortgage.ts, col ="grey",
       main = "30-Year Fixed Rate Mortgage Average in the United States", ylab = "Percent")
```
```{r}

```


From the plot of Median Sales Price of Houses Sold, we can see that there exists an obvious trend in the data. Therefore de-trending needs to be carried out to achieve stationarity.
```{r warning=FALSE}
# De-trending MSPUS Time Series
mspus_trend <- lm(mspus.ts ~ time(mspus.ts))
#summary(mspus_trend)

par(mfrow=c(2,1))
tsplot(mspus.ts, col ="darkorange",
       main = "Median Sales Price of Houses Sold for the United States",
       ylab = "USD")
abline(mspus_trend, col = "darkblue")
tsplot(resid(mspus_trend), col ="darkorange",
       main = "[De-trended] Median Sales Price of Houses Sold for the United States")

# check for Stationarity using (Augmented) Dickey-Fuller Test and KPSS Test
adf.test(resid(mspus_trend))
kpss.test(resid(mspus_trend))
```

Both ADF and KPSS tests conclude that the MSPUS de-trended series is non-stationary.
We will try differencing to achieve stationarity.

```{r warning=FALSE}
# Differencing MSPUS Time Series
tsplot(diff(mspus.ts))

adf.test(diff(mspus.ts))
kpss.test(diff(mspus.ts))
```

The series passes both these tests of stationarity. But the series exhibits an obvious heteroscedasticity where higher levels are associated with higher variation. A log-transformation is recommended.

```{r}
mspus.rs_log <- log(mspus.ts)
tsplot(diff(mspus.rs_log), main = "Return : Differencing after log transformation") # Return (Differencing after log transformation)
```

The BP Test for homoscedasticity fails for MSPUS, log(MSPUS), diff(MSPUS), diff(MSPUS, lag =4),  diff(log(MSPUS)), and diff(resid(mspus_trend)). Will try differencing with a lag of 4 after log transformation.

```{r warning=FALSE}
# Differencing with a lag of 4 after taking log
mspus.ts_log4 <- diff(log(mspus.ts), lag = 4)
tsplot(mspus.ts_log4)
adf.test(mspus.ts_log4)
kpss.test(mspus.ts_log4)
bptest(lm(mspus.ts_log4 ~ time(mspus.ts_log4)))

# Log transformation of the differencing at lag 4 is not carried out because it produces NaNs
```

Differencing with a lag of 4 after taking log passes ADF Test & KPSS tests of stationarity and BP-Test for homoscedasticity. Taking lag = 4 insted of the default lag = 1 while differencing seems to have helped in removing the seasonality associated with the MSPUS series.

```{r}
par(mfrow=c(3,2))
tsplot(log(mspus.ts), main = "Log Transformation") # Log Transformation
tsplot(diff(mspus.ts), main = "Differencing") # Differencing
tsplot(diff(log(mspus.ts)), main = "Return : Differencing after log transformation") # Return (Differencing after log transformation)
tsplot(diff(residuals(mspus_trend)), main = "Differencing after de-trending") # Differencing after de-trending
tsplot(diff(mspus.ts, lag = 4), main = "Differencing with lag = 4 (Annual)") # Differencing with a lag of 4
tsplot(diff(log(mspus.ts), lag = 4), main = "Differencing with lag = 4 after log transformation")
```

```{r}
# ACF of transformed outcome variable
acf(mspus.ts_log4, lag.max = 50)
```


```{r}
#msacsr_trend <- lm(msacsr.ts ~ time(msacsr.ts))
#summary(msacsr_trend)
tsplot(msacsr.ts, col ="mediumblue",
       main = "Monthly Supply of New Houses in the United States",
       ylab = "Month's Supply")
#abline(msacsr_trend, col = "red")
#tsplot(resid(msacsr_trend), col = "darkblue",
#       main = "[De-trended] Monthly Supply of New Houses in the United States")
adf.test(msacsr.ts)
kpss.test(msacsr.ts)
#bptest(lm(msacsr.ts ~ time(msacsr.ts)))
#bptest(lm(log(msacsr.ts) ~ time(msacsr.ts)))
```

Thus the MSACSR series is stationary.


```{r}
tsplot(houst.ts, col ="limegreen",
       main = "New Privately-Owned Housing Units Started : Total Units",
       ylab = "Thousands of Units")
adf.test(houst.ts)
kpss.test(houst.ts)
```

The ADF Test & KPSS tests confirm stationarity of HOUST.

```{r}
tsplot(rhorusq156n.ts, col ="red",
       main = "Homeownership Rate in the United States",
       ylab = "Percent")
adf.test(rhorusq156n.ts)
kpss.test(rhorusq156n.ts)

rhorusq156n.ts_diff <- diff(rhorusq156n.ts)
tsplot(rhorusq156n.ts_diff)
adf.test(rhorusq156n.ts_diff)
kpss.test(rhorusq156n.ts_diff)
bptest(lm(rhorusq156n.ts_diff ~ time(rhorusq156n.ts_diff)))
```

Thus Home Ownership series passes the tests for stationarity and homoscedasticity.

```{r}
tsplot(mortgage.ts, col ="grey",
       main = "30-Year Fixed Rate Mortgage Average in the United States", ylab = "Percent")
adf.test(mortgage.ts)
kpss.test(mortgage.ts)

mortgage.ts_diff <- diff(mortgage.ts)
tsplot(mortgage.ts_diff)
adf.test(mortgage.ts_diff)
kpss.test(mortgage.ts_diff)
bptest(lm(mortgage.ts_diff ~ time(mortgage.ts_diff)))
```


```{r}
# Clipping the series to bring them to the same length
msacsr.ts <- ts(MSACSR$MSACSR, start = c(1972, 2), frequency = 4, end = c(2023, 4))
houst.ts <- ts(HOUST$HOUST, start = c(1972, 2), frequency = 4, end = c(2023, 4))
mortgage.ts <- ts(MORTGAGE30US$MORTGAGE30US, start = c(1972, 1) , frequency = 4, end = c(2023, 4))
mortgage.ts_diff <- diff(mortgage.ts)
```


```{r}
trend <- time(mspus.ts_log4)
reg_model <- lm(mspus.ts_log4 ~ trend + msacsr.ts + houst.ts + mortgage.ts_diff)
summary(reg_model)
summary(aov(reg_model))
```

